A Mountaintop View Requires Minimal Sorting: \break An Instance Optimal Contour Tree Algorithm

### 2014-10-24 paint-contour-tree.tex
small changes

\newcommand{\build}{{\tt build}}
\newcommand{\col}{col}
\newcommand{\fcol}{full}
\newcommand{\cut}{{\tt cut}}
\newcommand{\cont}{\psi}
\newcommand{\cost}{\mathop{cost}}
\newcommand{\h}{att}
\newcommand{\lift}{{\tt lift}}
\newcommand{\mcol}{mcol}
\newcommand{\merge}{{\tt merge}}
\newcommand{\onestep}{{\tt piece}}
\newcommand{\paint}{{\tt paint}}
\newcommand{\pmax}{P_{\max}}
\newcommand{\rain}{{\tt rain}}
\newcommand{\reeb}{\cC}
\newcommand{\redH}{\widetilde{H}}
\newcommand{\rep}{rep}
\newcommand{\st}{K}
\newcommand{\surgery}{{\tt surgery}}
\newcommand{\touch}{T}
\newcommand{\update}{{\tt update}}
\newcommand{\wet}{{\tt wet}}

@@

There have also been some output sensitive results for computing the contour tree.  Let $t'$ be the number 
of critical points in the input mesh, which could potentially be much smaller than $n$.  In particular, 
Cole-McLaughlin and Pascucci \cite{pc-ectls-02} provide an $O(n+t\log n)$ time algorithm for the case of 
$3$ dimensional structured meshes.  \Ben{Update distinction between critical and component-critical}
Chiang \etal \cite{cllr-sooscctmp-05} improve upon this result, and provide an $O(N+t\log t)$ algorithm 
which works for structured or unstructured meshes in any dimension, which is provably optimal with respect 
to the output size but not the output structure, as discussed below.

@@

\subsection{Our Contribution}
\Ben{To be Written}

\paragraph{Organization.}

\begin{figure}[h]\centering
    \includegraphics[width=.18\linewidth]{figs/orderGray1}%
    \hspace{.8in}
    \includegraphics[width=.18\linewidth]{figs/orderGray2}
    \caption{On left, downward paint spilling only (each shade of gray represents a different color).  On right, flipping the direction of paint spilling.}
    \label{fig:order}
\end{figure}

@@

We detail the basic definitions about contour trees. Most of this is taken from Chapter 6 of Carr's thesis~\cite{Carr}.
Our input is a continuous PL function $f:\MM \mapsto \RR$, where $\MM$ is a simplicial complex in $\RR^d$. 
All simplices are isomorphic to a standard $d$-dimensional simplex, except for specially designated
\emph{boundary} simplices. \Sesh{Some mention of representation and data structures here.}
It is standard to assume
that $f$ is Morse (we explain later what this entails).

@@

Not that a contour is itself a simplicial complex of one dimension lower, and is represented (in our algorithms) as such.
We let $\delta, \eps$ denote infinitesimals. Let $B_\eps(x)$ denote a ball of radius $\eps$ around $x$.

@@

@@

\begin{definition} \label{def:bound} $f$ is \emph{boundary critical} if the following holds.
Consider a boundary simplex $F$. All vertices of $F$ have the same function value. Furthermore, all
neighbors of $F$ not in $F$ either have function values strictly greater or strictly less
than this value.
\end{definition}

@@

\begin{definition} \label{def:tree} The contour tree is the graph on vertex set $\cV$, where
edges are formed as follows. For every contour class that is created at $v_i$ and destroyed $v_j$,
there is an edge $(v_i,v_j)$. (Conventially, edges are directed from higher function value to lower.)
\end{definition}

(It is not immediately obvious that this graph is a tree. But it is equivalent to the Reeb graph
of $f:\MM \rightarrow \RR$, which is known to be a tree.) We denote the contour tree of $\MM$
by $\reeb(\MM,f)$. The corresponding node and edge sets are denoted as $\cV(\cdot)$ and $\cE(\cdot)$.

We use a key theorem from~\cite{Carr} (Theorem 6.6) that maps paths in $\cC(\MM)$ and $\MM$.

@@

{\bf The cutting operation:} We define a ``cut" operation on $f:\MM \rightarrow \RR$ that cuts along a regular contour to create
a new simplicial complex with an added boundary. Given contour $\phi$, this is constructing the simplicial complex $\MM \setminus \phi$. 
We will always enforce the condition that $\phi$ never passes through a vertex of $\MM$.
Again, we use $\eps$ for an infinitesimally small value. We denote $\phi^+$ (resp. $\phi^-$) to be
the contour at value $f(\phi^+) + \eps$ (resp. $f(\phi^-) - \eps$), which is at distance $\eps$
from $\phi$.

An $h$-contour is achieved by intersecting $\MM$ with the hyperplane $x_{d+1} = h$. (Think of the $d+1$-dimension
as height.) Given some point $x$
on an $h$-contour $\phi$ (with some simplex that $x$ is present in), we can walk along $\MM$ from $x$ and determine $\phi$.
We can ``cut" along $\phi$ to get a new (possibly) disconnected simplicial complex $\MM'$. This is achieved
by splitting every facet $F$ that $\phi$ intersects into an ``upper" facet and ``lower" facet. Algorithmically,
we cut facet $F$ with $\phi^+$ and take everything above $\phi^+$ in $F$ to make the upper facet. Analogously, we cut with $\phi^-$ to get the lower facet.
The facets are then triangulated to ensure that they are all simplices.
Given that $\phi$ cannot cut a boundary and all non-boundary facets
have constant size, we omit the algorithmic description for this process. We only note that this
creates the two new boundaries $\phi^+$ and $\phi^-$, and we maintain the property of constant $f$-value at a boundary.
This new simplicial complex is denoted by $\cut(\phi,\MM)$.
This can be constructed in time linear in $|\phi|$.

@@

\begin{lemma} \label{lem:cut} Consider a regular contour $\phi$ contain in contour class (edge of $\reeb(\MM))$
$(u,v)$ and denote $\MM' = \cut(\phi,\MM)$. Then $\cV(\reeb(\MM')) = \{\phi^+,\phi^-\} \cup (\cV(\MM) \setminus \phi)$
and $\cE(\reeb(\MM')) = \{(u,\phi^+),(\phi^-,v)\} \cup (\cE(\MM) \setminus (u,v))$.
\end{lemma}

@@

Any contour class in $\MM'$ (edge in $\cC(\MM')$) that does not involve $\phi^+$ or $\phi^-$ is also
a contour class in $\MM$. Furthermore, a maximal contour class satisfying these properties is also
maximal in $\MM$. So all edges of $\cC(\MM')$ that do not involve $\phi^+$ or $\phi^-$ are edges of $\cC(\MM)$.
Analogously, every edge of $\cC(\MM)$ not involving $\phi$ is an edge of $\cC(\MM)$.

@@

\begin{proof} Denote (as in \Lem{cut}) the edge containing $\phi$ to be $(u,v)$. Suppose for contradiction that there is a path between vertices $u$ and $v$
in $\MM' = \cut(\phi,\MM)$. By \Thm{carr-path}, there is a path in $\cC(\MM')$ between $u$ and $v$. Since $\phi^+$ and $\phi^-$
are leaves in $\cC(\MM')$, this path obviously cannot use edges incident to them. By \Lem{cut},
all the edges of this path are in $\cE(\cC(\MM)) \setminus (u,v)$. So we get a cycle in $\cC(\MM)$. Contradiction.
To show that there are exactly two connected components in $\cut(\phi,\MM)$, it suffices
to see that $\cC(\MM')$ has two connected components (by \Lem{cut}) and apply \Thm{carr-path}.
\end{proof}

@@

\begin{definition} \label{def:dom} A simplicial complex is \emph{minimum dominant} if there exists
a minimum $x$ such that every non-minimal \emph{vertex} in the manifold has decreasing path to $x$.
Analogously define \emph{maximum dominant}. 
\end{definition}

@@

\begin{definition} \label{def:wet} Fix $\MM$ and $x \in \MM$. The set of points $y \in \MM$ such that there is a non-ascending path from $x$ to $y$
is denoted by $\wet(x,\MM)$ (which is in turn represented as a simplicial complex). A point $z$ is at the \emph{interface} of $\wet(x,\MM)$ if every neighborhood of $z$
has non-trivial intersection with $\wet(x,\MM)$.
\end{definition}

@@

\begin{proof} If $y \in \wet(x,\MM)$, all the points in any contour containing $y$ are also in $\wet(x,\MM)$.
 (Follow the non-ascending path to from $x$ to $y$ and then walk along the contour.) The converse is also true,
 so $\wet(x,\MM)$ contains entire contours.

Let $\eps, \delta$ be sufficiently small as usual. Fix some $y$ at the interface.
Note that $y \in \wet(x,\MM)$. (Otherwise, $B_\eps(y)$ is dry.)
The points in $B_\eps(y)$ that lie below $y$ have a descending path from $y$ and hence must be wet.
There must be a dry point in $B_\eps(y)$ that is above $y$, and hence,
there exists a dry, regular $(f(y)+\delta)$-contour $\phi$ intersecting $B_\eps(y)$.

Suppose $y$ is in the interface. Consider the set $\Gamma_y$ of points connected to $y$
by a path with all points at height $f(y)$. (This is a set of $f(y)$-contours.) Suppose for contradiction
that $\forall z \in \Gamma_y$, $z$ has up-degree $1$. Consider the non-ascending path from $x$ to $y$ and let $z$
be the first point of $\Gamma_y$ encountered. There exists a wet, regular $(f(y) + \delta)$-contour $\psi$ 
intersecting $B_\eps(z)$. Now, walk from $z$ to $y$ along $\Gamma_y$. If all points $w$ in this walk
have up-degree $1$ (ref. to \Def{deg}), then $\psi$ is the unique $f(y)+\delta$-contour
intersecting $B_\eps(w)$. This would imply that $\phi = \psi$, contradicting the fact that $\phi$ is wet
and $\psi$ is dry.

Therefore, there must exist a join vertex $w$ in $\Gamma_y$, such that $B_\eps(w)$ intersects
both $\phi$ and $\psi$. As $\delta, \eps \rightarrow 0^+$, the limit of $\phi$ (which is dry)
lies in $\Gamma_y$ (which is wet). Hence, this gives the interface, which is a contour containing
the join $w$.
\end{proof}

@@

Note that $\wet(x,\MM)$ (and its interface) can be computed in time linear in the size of the wet simplicial complex.
We perform a non-ascending walk from $x$. Any facet $F$ of $\MM$ encountered has some (if not all) portion of in $\wet(x,\MM)$.
This portion is determined by cutting $F$ along the interface. Since the interface is a contour, this is equivalent
to cutting $F$ by a hyperplane. All these operations can be performed to output $\wet(x,\MM)$ in time linear in $|\wet(x,\MM)|$.

@@

We define a simple \lift{} operation on the interface contours. Consider such a contour $\phi$ containing
a merge vertex $y$. Take any dry increasing edge incident to $y$, and pick the point $z$ on this edge at height
$f(y) + \delta$. Let $\lift(\phi)$ be the unique contour through regular point $z$. Note that $\lift(\phi)$ is dry.

@@

We describe the main partitioning procedure that cuts a simplicial complex $\NN$ into extremum
dominant complices. It takes an additional input of a maximum $x$. To initialize,
we begin with $\NN$ set to $\MM$ and $x$ as an arbitrary maximum. One of the critical aspects
of the procedure is that the rain alternately flows downwards and upwards. In other words, when
we start, rain flows downwards. In each recursive call, the direction of rain is switched to the 
opposite direction. While this is way of implementing $\rain$, it is conceptually easier
to think of \emph{inverting} a complex $\NN'$ when a recursive call is made. Inversion
is easily achieved by just negating the height values. (From a running time standpoint, it suffices
to maintain a single bit associated with $\NN'$ that determines whether heights are inverted or not.)
We can now let rain flow downwards, as it usually does in our world.

@@

\begin{lemma}\label{lem:new-verts}
A facet $F \in \MM$ is cut at most once during the $\rain(x,\MM)$.
\end{lemma}
\begin{proof} All notation here follows that in the pseudocode of $\rain$.
Firstly, by \Thm{jordan}, all the pieces of which $\rain$ is invoked are disjoint.
Secondly, all recursive calls are made on dry complices.

Consider the first time that $F$ is cut, say, during the call to $\rain(x,\NN)$.
Specially, say this happens when $\cut(\phi'_i,\NN_i)$ is constructed. 
So $F$ is cut by some hyperplane into lower and upper portions (which are then triangulated).
Since the cut is infinitesimally above an interface, the lower portion is basically wet.
It can never be part of $\LL_1, \LL_2, \ldots$, which go into recursive calls. This portion
is in $\NN_{k+1}$. The upper portion/facet (call it $U$) is in $\LL_i$. Note that the lower
boundary of $U$ is in the boundary $B_i$. Since a recursive call is made to $\rain(B_i,\LL_i)$
(and $\LL_i$ is inverted), $U$ becomes wet. Hence, $U$ will not be cut subsequently.
All in all, $F$ is cut at most once.
\ignore{
As the number of edges is $O(n)$, arguing that there is at most one boundary vertex created along any given edge will imply the claim.
Consider some edge $e$ and take the first time a boundary vertex is created in $e$.
This happens when some contour $\phi$ intersects $e$. On applying $\cut(\NN,\lift(\phi))$,
consider the dry manifold $\LL$ and wet manifold $\MM$ constructed. The edge $e$ is split into two parts, the dry and wet part.
The wet part is in $\MM$ and will never be involved in any recursive call.

@@

\begin{proof} The operations performed are the raining and cutting. We can bound this by the total
complexities of the wet complices and the new boundaries created. No facet is every wet twice,
and by \Lem{new-verts}, the total number of facets that every appear is $O(|\MM|)$. This also 
bounds the complexities of the boundaries created.
\end{proof}

@@

Now we come to the main algorithmic contribution. We design a new algorithm for computing contour
trees of a minimum dominant simplicial complice $\MM$. As part of the initialization, we determine
all critical points in $\MM$. For any vertex $v$, we can look at $f(v) \pm \delta$-contours 
in $f|_{B_\eps(v)}$ to determine the up and down degrees. This determines the criticality (and type)
of $v$. This requires $O(|\MM|)$ time, since every facet is looked at proportional to the number of its
vertices.

{\bf Painting:} The central tool is a notion 
of \emph{painting} of $\MM$. Initially associate a color with each maximum. Denoting the set of maxima is $X$,
this is an arbitrary bijection $\chi:X \mapsto [|X|]$. Imagine there being a large
can of paint of color $\chi(x)$ at maxima $x$. We will spill different paint from each maximum and watch it flow down.
This is analogous to the raining in the previous section, but paint is a much more viscous liquid.
\emph{So paint only flows down edges, and it does not color the interior of facets.} Furthermore, paints
do not mix, so eventually, every vertex and edge of $\MM$ gets a unique color. This process (and indeed the entire algorithm)
works purely on the 1-skeleton of $\MM$, which is just a graph.
%
\begin{definition} \label{def:paint} Let the 1-skeleton of $\MM$ be denoted $G(\MM) = (\cV,E)$.
A  \emph{painting} of $\MM$ is a map $\chi:X \cup E \mapsto [|X|]$ with the following properties. 
\begin{asparaitem}
	\item The restriction $\chi:X \mapsto [|X|]$ is a bijection.
	\item Consider an edge $e$. There exists a descending path from some maximum $x$ to $e$
	consisting of edges in $E$, such that all edges along this path have the same color as $x$. (The maximum $x$
	is well-defined to be the unique maximum with the same color as $e$.)
\end{asparaitem}
\end{definition}
%
Note that paintings are not unique. It is straightforward to construct a painting of $\MM$ in $O(|\MM|)$ time, using a descending BFS from each maximum that does not explore previously colored edges.

{\bf Colors of vertices:} Note that a painting only colors edges, not vertices. We associate certain
sets of colors with vertices.

\begin{definition} \label{def:color-set} Fix a painting $\chi$, and a color $c$.
A vertex $v$ is \emph{touched by $c$} if $x$ is incident to a $c$-colored
edge with $v$ at the lower endpoint. For $v$, $\col(v)$ is the set of colors that touch $v$.
\end{definition}

Abusing notation, we say a contour $\phi$ is of color $c$ if all edges that intersect $\phi$ have color $c$.

\begin{definition} \label{def:color-full} Consider vertex $v$, and let $\phi$ be a $(f(v)+\delta)$-contour in
$f|_{B_\eps(v)}$. If $\phi$ is monochromatic of color $c$, then $v$ is \emph{fully touched by $c$}. The set of colors
that fully touch $v$ is denoted $\fcol(v)$.
\end{definition}

Note that $\fcol(v)$ has size at most $2$, by the Morse condition. Note that $\col(v)$ and $\fcol(v)$ can
be initially determined in $O(|\MM|)$ time, by a simple linear scan.

>>>>

\newcommand{\build}{{\tt build}}
\newcommand{\col}{col}
\newcommand{\cut}{{\tt cut}}
\newcommand{\cont}{\psi}
\newcommand{\cost}{\mathop{cost}}
\newcommand{\h}{high}
\newcommand{\lift}{{\tt lift}}
\newcommand{\mcol}{mcol}
\newcommand{\merge}{{\tt merge}}
\newcommand{\onestep}{{\tt piece}}
\newcommand{\paint}{{\tt paint}}
\newcommand{\pmax}{P_{\max}}
\newcommand{\rain}{{\tt rain}}
\newcommand{\reeb}{\cC}
\newcommand{\redH}{\widetilde{H}}
\newcommand{\rep}{rep}
\newcommand{\st}{K}
\newcommand{\surgery}{{\tt surgery}}
\newcommand{\touch}{T}
\newcommand{\update}{{\tt update}}
\newcommand{\wet}{{\tt wet}}

@@

There have also been some output sensitive results for computing the contour tree.  Let $t'$ be the number 
of critical points in the input mesh, which could potentially be much smaller than $n$.  
Cole-McLaughlin and Pascucci \cite{pc-ectls-02} provided an $O(n+t'\log n)$ time algorithm for the case of 
$3$-dimensional structured meshes (note $N$ does not show up in the running time as the mesh is structured).
Chiang \etal \cite{cllr-sooscctmp-05} improve upon this result, and provide an $O(N+t\log t)$ algorithm 
which works for structured or unstructured meshes in any dimension, where $t$ is the number of component 
critical vertices.  This is provably optimal with respect to the output size but not the output structure, 
as discussed below.

@@

\subsection{Overview and Contribution}

\paragraph{Algorithm Overview.}
At a high level, the technique is rather simple.  Break the input manifold into nicely 
behaved pieces, compute the contour tree on each piece, and then stitch together the computed 
contour trees according to the cuts we made to break up the manifold.  Below we informally 
describe how this is accomplished.
\begin{compactitem}
 \item {\bf Raining to partition.}  To start with, pick an arbitrary maxima.  Now imagine 
 ``raining'' water down on the manifold from this maxima, i.e. mark all points which are reachable 
 by non-ascending paths from the maxima.  Let the boundary between wet and dry portions of the 
 manifold be called the interface (which may have different connected components).  
 Remove the wet submanifold, and invert the remaining pieces
 of the original manifold.  Now recursively rain in each one of these pieces, but rather than 
 picking an arbitrary maxima, start from the portion of the interface that cut that piece off.

 The benefit of this partitioning is that each resulting piece 
 is maxima dominant, i.e. there is a single maxima that has a non-ascending path to each point.  
 This implies the contour of this submanifold is the same as its split tree, or in other words 
 its contour tree is a rooted binary tree.

 %
 \item {\bf Painting trees.}  For each maxima dominant piece we want to compute the split 
 tree without sorting all the vertices.  For conceptual ease lets first flip the submanifold 
 (i.e., it is now minima dominant and we compute the join tree).  Now arbitrarily pick a maxima 
 and spill paint from it, specifically color all edges and vertices of the mesh that are reachable 
 from the maxima by descending paths.  Now go to each other maxima and spill a different color, 
 but be aware that paints do not mix, i.e. if we reach a vertex that was previosly explored 
 by another color then we do not explore it.

 This procedure partitions the set of critical points into color classes.  Clearly each color 
 class is connected in the manifold.  Suppose each class was also connected in the join tree.  
 In this case we partitioned the join tree into a set of descending paths, each of which 
 can be constructed by sorting the critical points in a class.  Roughly speaking, 
 these paths can then be joined together by what amounts to doing a post-order traversal of the tree.
 However, we may not be so lucky and the color classes may not be connected in the 
 tree.  This turns out not to be a problem, but it requires proper data-structure use. 
%  
%  Now suppose we were lucky and the spilling of paint in the manifold corresponded with a paint 
%  spilling in the join tree, i.e. the color classes of the critical points partition the join 
%  tree into disjoint descending paths, each containing one leaf.   In this case 
%  we have partitioned the tree into branches, and so the tree can be formed by gluing these 
%  branches together.  Intuitively (i.e. not exactly true), if two branches glue together then 
%  it means there is a join vertex which got colored by the respective colors of each branch 
%  (only one of these colors actually explored the vertex in the mesh and this is the ``main'' 
%  branch you are gluing into).  Therefore, the tree can be constructed by starting at the leaf 
%  of the trunk (i.e. the branch with the global minima) and recursively building the subtrees 
%  while walking towards the root.  In other words we are building the tree by doing a post-order traversal.
%  Unfortunately, we may not be so lucky and the color classes may not appear contiguously in 
%  the tree.  In this case we can still build the tree in this leaf first fashion, but doing so 
%  efficiently requires the use of data-structures.
\end{compactitem}

\paragraph{Lower Bound Overview.}
\Ben{To be Written}

\paragraph{Our Contribution.}
Our contribution can be broken into three main aspects.  The first two are algorithmic.  Namely, 
first we describe a new linear time manifold partitioning scheme which breaks the manifold into 
``extremum'' dominant submanifolds.  
Second, we describe how to compute join trees optimally with respect to shape of the tree.
As the contour tree of each submanifold is (roughly) equal to its join tree, this leads to an 
optimal algorithm for the contour tree of the input manifold.  
Therefore, our third and final contribution is proving the optimality of this result, which 
requires a non-trivial amount of work, and uses techniques we believe to be of independent interest.

\paragraph{Organization.}
\Ben{To be Written}

@@

We detail the basic definitions about contour trees. The following is based on Chapter 6 of Carr's thesis \cite{c-tmi-04}.
Our input is a continuous PL function $f:\MM \mapsto \RR$, where $\MM$ is a fully triangulated simplicial complex in $\RR^d$. 
%All simplices are isomorphic to a standard $d$-dimensional simplex, except for specially designated \emph{boundary} simplices. 
\Sesh{Some mention of representation and data structures here.}
We use the standard assumption that $f$ is Morse (we explain later what this entails).

@@

Note that a contour is itself a simplicial complex of one dimension lower, and is represented (in our algorithms) as such.
We let $\delta, \eps$ denote infinitesimals. Let $B_\eps(x)$ denote a ball of radius $\eps$ around $x$.

@@

\begin{remark}
 For two manifolds, at each saddle point either two contours join together or split apart, and hence the terminology.  In higher dimensions this may not be the case, however, it always holds that at a join or split vertex the topology of the level set changes, and hence such vertices are still 'critical'.  Note that determining which category each vertex falls into according to the above definition can be determined for all vertices simultaneously in linear time, as it is a local neighborhood condition.
\end{remark}

@@

\begin{definition} \label{def:bound} $f$ is \emph{boundary critical} if the following holds.
Consider a boundary simplex $F$. All vertices of $F$ have the same function value. Furthermore, all
neighbors of $F$ not in $F$ either have all function values strictly greater or strictly less
than this value.
\end{definition}

\begin{remark}
Note that by assuming that our manifold is boundary critical, the vertices on a given boundary are either all maxima or all minima.  
Since all such vertices on a boundary component are in the same contour, we will abuse notation and refer to this entire set of vertices as a maxima (resp. minima).
In particular, we do not view this as a violation of the Morse condition on $f$, which requires distinct heights for critical points.
\end{remark}

@@

\begin{definition} \label{def:tree} The contour tree is the graph on vertex set $\cV$, where
edges are formed as follows. For every contour class that is created at $v_i$ and destroyed $v_j$,
there is an edge $(v_i,v_j)$. (Conventionally, edges are directed from higher function value to lower.)
\end{definition}

We denote the contour tree of $\MM$
by $\reeb(\MM) = \reeb(\MM,f)$. The corresponding node and edge sets are denoted as $\cV(\cdot)$ and $\cE(\cdot)$.
%
It is not immediately obvious that this graph is a tree, but the above definition is equivalent to that of the Reeb graph of $f:\MM \rightarrow \RR$ on $\MM$, and this is known to be a tree for PL functions defined over PL manifolds.

\begin{remark}
Note that the contour tree as defined above contains vertices for all critical points, 
which differs slightly from the usual definition which only contains critical points corresponding to a change in the number of contours.  Specifically, if a vertex in the tree we defined has degree $2$ then only the topology of a contour is changing rather than the number of contour.  As such the standard contour tree can be computed from ours by a linear time pass to contract maximal degree 2 vertex paths.

For simplicity we will also make the standard assumption that vertices in the contour 
tree have degree at most $3$, i.e. there are no multi-saddles.
\end{remark}

We use a key theorem from \cite{c-tmi-04} (Theorem 6.6) that maps paths in $\cC(\MM)$ and $\MM$.

@@

{\bf The cutting operation:} We define a ``cut" operation on $f:\MM \rightarrow \RR$ that cuts along a regular contour to create
a new simplicial complex with an added boundary. Given contour $\phi$, this is constructing the simplicial complex $\MM \setminus \phi$. 
We will always enforce the condition that $\phi$ never passes through a vertex of $\MM$.
Again, we use $\eps$ for an infinitesimally small value. We denote $\phi^+$ (resp. $\phi^-$) to be
the contour at value $f(\phi) + \eps$ (resp. $f(\phi) - \eps$), which is at distance $\eps$
from $\phi$.

An $h$-contour is achieved by intersecting $\MM$ with the hyperplane $x_{d+1} = h$ and taking a connected component. (Think of the $d+1$-dimension
as height.) Given some point $x$
on an $h$-contour $\phi$ (with some simplex that $x$ is present in), we can walk along $\MM$ from $x$ and determine $\phi$.
We can ``cut" along $\phi$ to get a new (possibly) disconnected simplicial complex $\MM'$. This is achieved
by splitting every facet $F$ that $\phi$ intersects into an ``upper" facet and ``lower" facet. Algorithmically,
we cut facet $F$ with $\phi^+$ and take everything above $\phi^+$ in $F$ to make the upper facet. Analogously, we cut with $\phi^-$ to get the lower facet.
The facets are then triangulated to ensure that they are all simplices.
Given that $\phi$ cannot cut a boundary and all non-boundary facets
have constant size, we omit the algorithmic description for this process. We only note that this
creates the two new boundaries $\phi^+$ and $\phi^-$, and we maintain the property of constant $f$-value at a boundary.
This new simplicial complex is denoted by $\cut(\phi,\MM)$.
This can be constructed in time linear in $|\phi|$.

@@

\begin{lemma} \label{lem:cut} Consider a regular contour $\phi$ contained in a contour class (of an edge of $\reeb(\MM))$
$(u,v)$ and denote $\MM' = \cut(\phi,\MM)$. Then $\cV(\reeb(\MM')) = \{\phi^+,\phi^-\} \cup \cV(\MM)$
and $\cE(\reeb(\MM')) = \{(u,\phi^+),(\phi^-,v)\} \cup (\cE(\MM) \setminus (u,v))$.
\end{lemma}

@@

Any contour class in $\MM'$ (edge in $\cC(\MM')$) that does not involve $\phi^+$ or $\phi^-$ is also
a contour class in $\MM$. Furthermore, a maximal contour class satisfying these properties is also
maximal in $\MM$. So all edges of $\cC(\MM')$ that do not involve $\phi^+$ or $\phi^-$ are edges of $\cC(\MM)$.
Analogously, every edge of $\cC(\MM)$ not involving $\phi$ is an edge of $\cC(\MM')$.

@@

\begin{proof} Denote (as in \Lem{cut}) the edge containing $\phi$ to be $(u,v)$. Suppose for contradiction that there is a path between vertices $u$ and $v$
in $\MM' = \cut(\phi,\MM)$. By \Thm{carr-path}, there is a path in $\cC(\MM')$ between $u$ and $v$. Since $\phi^+$ and $\phi^-$
are leaves in $\cC(\MM')$, this path obviously cannot use edges incident to them. By \Lem{cut},
all the edges of this path are in $\cE(\cC(\MM)) \setminus (u,v)$. So we get a cycle in $\cC(\MM)$, a contradiction.
To show that there are exactly two connected components in $\cut(\phi,\MM)$, it suffices
to see that $\cC(\MM')$ has two connected components (by \Lem{cut}) and apply \Thm{carr-path}.
\end{proof}

@@

\begin{definition} \label{def:dom} A simplicial complex is \emph{minimum dominant} if there exists
a minimum $x$ such that every non-minimal \emph{vertex} in the manifold has a decreasing path to $x$.
Analogously define \emph{maximum dominant}. 
\end{definition}

@@

\begin{definition} \label{def:wet} Fix $\MM$ and $x \in \MM$. The set of points $y \in \MM$ such that there is a non-ascending path from $x$ to $y$
is denoted by $\wet(x,\MM)$ (which is in turn is represented as a simplicial complex). A point $z$ is at the \emph{interface} of $\wet(x,\MM)$ if every neighborhood of $z$
has non-trivial intersection with $\wet(x,\MM)$.
\end{definition}

@@

\begin{proof} If $p \in \wet(x,\MM)$, all the points in any contour containing $p$ are also in $\wet(x,\MM)$.
 (Follow the non-ascending path to from $x$ to $p$ and then walk along the contour.) The converse is also true,
 so $\wet(x,\MM)$ contains entire contours.

Let $\eps, \delta$ be sufficiently small as usual. Fix some $y$ at the interface.
Note that $y \in \wet(x,\MM)$. (Otherwise, $B_\eps(y)$ is dry.)
The points in $B_\eps(y)$ that lie below $y$ have a descending path from $y$ and hence must be wet.
There must also be a dry point in $B_\eps(y)$ that is above $y$, and hence,
there exists a dry, regular $(f(y)+\delta)$-contour $\phi$ intersecting $B_\eps(y)$.

Let $\Gamma_y$ be the contour containing $y$.  Suppose for contradiction
that $\forall p \in \Gamma_y$, $p$ has up-degree $1$ (see \Def{deg}). Consider the non-ascending path from $x$ to $y$ and let $z$
be the first point of $\Gamma_y$ encountered. There exists a wet, regular $(f(y) + \delta)$-contour $\psi$ 
intersecting $B_\eps(z)$. Now, walk from $z$ to $y$ along $\Gamma_y$. If all points $w$ in this walk
have up-degree $1$, then $\psi$ is the unique $f(y)+\delta$-contour
intersecting $B_\eps(w)$. This would imply that $\phi = \psi$, contradicting the fact that $\psi$ is wet
and $\phi$ is dry.

Therefore, there must exist a join vertex $w$ in $\Gamma_y$, such that $B_\eps(w)$ intersects
both $\phi$ and $\psi$. As $\delta, \eps \rightarrow 0^+$, the limit of $\phi$ (which is dry)
lies in $\Gamma_y$ (which is wet). 
Hence this limit gives a contour of the interface containing the join $w$.
\end{proof}

@@

Note that $\wet(x,\MM)$ (and its interface) can be computed in time linear in the size of the wet simplicial complex.
We perform a non-ascending search from $x$. Any facet $F$ of $\MM$ encountered is partially (if not entirely) in $\wet(x,\MM)$.
This portion is determined by cutting $F$ along the interface. Since the interface is a contour, this is equivalent
to cutting $F$ by a hyperplane. All these operations can be performed to output $\wet(x,\MM)$ in time linear in $|\wet(x,\MM)|$.

@@

We define a simple \lift{} operation on the interface contours. Consider such a contour $\phi$ containing
a merge vertex $y$. Take any dry increasing edge incident to $y$, and pick the point $z$ on this edge at height
$f(y) + \delta$. Let $\lift(\phi)$ be the unique contour through the regular point $z$. Note that $\lift(\phi)$ is dry.

@@

\begin{figure}[h]\centering
    \includegraphics[width=.18\linewidth,height=.35\linewidth]{figs/orderGray1}%
    \hspace{.8in}
    \includegraphics[width=.18\linewidth,height=.35\linewidth]{figs/orderGray2}
    \caption{On left, downward paint spilling only (each shade of gray represents a different color), producing a grid.  On right, flipping the direction of paint spilling.}
    \label{fig:order}
\end{figure}

We describe the main partitioning procedure that cuts a simplicial complex $\NN$ into extremum
dominant complices. It takes an additional input of a maximum $x$. To initialize,
we begin with $\NN$ set to $\MM$ and $x$ as an arbitrary maximum. One of the critical aspects
of the procedure is that the rain alternately flows downwards and upwards, 
since otherwise faces may be cut a super constant number of times (see \Fig{order}).
In other words, when
we start, rain flows downwards. In each recursive call, the direction of rain is switched to the 
opposite direction. While one can implement $\rain$ this way, it is conceptually easier
to think of \emph{inverting} a complex $\NN'$ when a recursive call is made. Inversion
is easily achieved by just negating the height values. (From a running time standpoint, it suffices
to maintain a single bit associated with $\NN'$ that determines whether heights are inverted or not.)
We can now let rain flow downwards, as it usually does in our world.

@@

\begin{lemma}\label{lem:new-verts}
A facet $F \in \MM$ is cut at most once during $\rain(x,\MM)$.
\end{lemma}
\begin{proof} All notation here follows that in the pseudocode of $\rain$.
First, by \Thm{jordan}, all the pieces on which $\rain$ is invoked are disjoint.
Second, all recursive calls are made on dry complices.

Consider the first time that $F$ is cut, say, during the call to $\rain(x,\NN)$.
Specifically, say this happens when $\cut(\phi'_i,\NN_i)$ is constructed. 
So $F$ is cut by some hyperplane into lower and upper portions (which are then triangulated).
Since the cut is infinitesimally above an interface, the lower portion is basically wet.
It can never be part of $\LL_1, \LL_2, \ldots$, which go into recursive calls. This portion
is in $\NN_{k+1}$. The upper portion/facet (call it $U$) is in $\LL_i$. Note that the lower
boundary of $U$ is in the boundary $B_i$. Since a recursive call is made to $\rain(B_i,\LL_i)$
(and $\LL_i$ is inverted), $U$ becomes wet. Hence, $U$ will not be cut subsequently.
All in all, $F$ is cut at most once.
\ignore{
As the number of edges is $O(n)$, arguing that there is at most one boundary vertex created along any given edge will imply the claim.
Consider some edge $e$ and take the first time a boundary vertex is created in $e$.
This happens when some contour $\phi$ intersects $e$. On applying $\cut(\NN,\lift(\phi))$,
consider the dry manifold $\LL$ and wet manifold $\MM$ constructed. The edge $e$ is split into two parts, the dry and wet part.
The wet part is in $\MM$ and will never be involved in any recursive call.

@@

\begin{proof} The operations performed are the raining and cutting. We can bound this by the total
complexities of the wet complices and the new boundaries created. No facet is ever wet twice,
and by \Lem{new-verts}, the total number of facets that every appear is $O(|\MM|)$. This also 
bounds the complexities of the boundaries created.
\end{proof}

@@

We begin by associating a color with each maximum. If the set of maxima is $S$,
this is simply an arbitrary bijection $\chi:S \mapsto [|S|]$. Imagine there being a large
can of paint of color $\chi(y)$ at maxima $y$. We will spill different paint from each maximum and watch it flow down.
This is analogous to the raining in the previous section, but paint is a much more viscous liquid.
\emph{So paint only flows down edges, and it does not color the interior of faces.} Furthermore, paints
do not mix, so eventually, every vertex and edge of $\MM$ gets a unique color.
%
\begin{definition} \label{def:paint} A  \emph{painting} of $\MM$ is a map $\chi:S \cup E \mapsto [|S|]$
with the following properties. (Remember that $G(\MM) = (V,E)$.)
\begin{asparaitem}
	\item The restriction $\chi:S \mapsto [|S|]$ is a bijection.
	\item Consider an edge $e$. There exists a descending path from some maximum $x$ to $e$
	consisting of edges in $E$, such that all edges along this path have the same color as $x$. (The maximum $x$
	is well-defined to be the unique maximum with the same color as $e$.)
\end{asparaitem}
\end{definition}
%
Note that paintings are not unique. It is straightforward to construct a painting of $\MM$ in $O(n)$ time, using a descending BFS from each maximum that does not explore previously colored edges.
